Lab 2 Final Report: COVID-19 Case Rate vs Population Demographics and Mask Policy

### 2020-12-06 Jackson_Liu_Temlock_Zhang_Lab2_Draft_Report.Rmd
Updated RMD and re-knit per meeting discussions

The research question aims to measure the COVID-19 case rate per 100,000 residents within each state based on the make-up of the population, as well as the policy decisions that were or were not put into place in order to combat the rise of COVID-19 cases. In this sense, the aim is to examine how dependent the COVID-19 case rate is on that of which cannot be controlled (i.e., population demographics), as well as that of which can be controlled to a certain degree (i.e., implementation of policies to attempt to combat case rate) by each state. With this information, one could suggest whether or not the proliferation of COVID-19 within a state seems to be related to either or both controllable and uncontrollable variables, and to the level at which the effect of these variables differ help. The modeling goal of this research question will be one of description, and will be broken up into three phases of investigation.

The first and primary phase of investigation involves how the COVID-19 case rate is affected by key population demographics. For this, the key variables are COVID-19 case rate per 100,000 residents and the percentage population distribution of seniors (defined as 65 years old or older) within a state. The case rate per 100,000 was selected as the key output/dependent variable as it provides a standardized measure of the spread of COVID-19 across the states, taking into account the absolute population of the states. The senior percentage was selected as the key input/independent variable given that guidance had been released by the Centers for Disease Control and Prevention (CDC) that seniors belong to the age category for those at higher risk of COVID-19, and thus are a well studied and documented group. This is likely due to the both the fact that seniors in this age group are more likely or had greater access to be tested given their categorization and the fact that they would have a greater likelihood of exhibiting detectable COVID-19 related symptoms that would prompt them to get tested. An additional rationale for measuring the senior percentage is that they may also be more likely to contract COVID-19 at lower viral loads, making them more susceptible to the virus. Conversely, this may result in seniors being more wary of the threat of COVID-19, and taking additional precautions to prevent infection relative to other age categories, such as limiting social interactions and taking more preventative measures in terms of hygiene. These key variables will provide an initial understanding of the relationship between the spread of the virus and population demographics.

Following this, in the second phase of investigation the analysis considers other variables of state-level population that are considered factors of susceptibility, namely the rate of poverty (defined as the percent of individuals living under the federal poverty line in 2018) and the rate of unemployment (defined as the percent unemployment in 2018). Those living below the poverty line may have less access/are unable to afford to preventative controls such as sanitation products and masks that help prevent the spread of the virus, while those who are designated as unemployed may not have the flexibility of sheltering at home as well as access to the aforementioned preventative controls. Alternatively, the likelihood of reduced mobility for those living under the poverty line and/or who are unemployed may also play a role in the relationship with case rate. As mentioned, these variables operationalize population metrics that may lead to greater case rates, and are already standardized as rates to account for varying absolute populations across states.

The tertiary and final phase of investigation includes a variable that measures the implementation of policy as a response to COVID-19, and to understand the added effect of this variable in conjunction with the first phase of investigation that is state population metrics. Specifically, there is a focus on a policy that mandates the wearing of masks within the state. Through this, the aim is to analyze whether the implementation of a mask-related policy will have an effect on case rate, as well as the degree of this effect relative to population metrics. To operationalize this policy and simplify the measurements, the models only consider whether or not this policy was implemented through transformed indicator variables (1 = implemented, 0 = not implemented). As a result, factors of when and for how long the policy was implemented will be lost. Although this loss of information may fail to capture the impact of length of time of a policy on case rate (the policy may take time to be adopted/show meaningful efficacy), as there is no time-series data for case rate included in the data set, it was adjudged to be incongruent with the analysis.

Additionally, the final phase of investigation continues to expand on the examination of state-level population demographics via the population density variable (defined as the population density in square miles, population/square miles). A key factor to the spread of COVID-19 is the idea of social distancing, where the virus is considered to be more likely to spread when people are close in proximity. Thus, population density is measured as a conduit to indicate the level of proximity within each state and will operationalize the concept of social distancing.

@@

Secondly, note that there are many internal and external aspects of the selected variables that have not been included within the models. Just addressing the internal information that is lost, it can be seen that some of the information is not captured given the methods of operationalization discussed previously. Among others, the loss of information on the date of implementation within the mask policy variable strips out any contextual knowledge regarding the length of policy implementation, and precludes the identification of how long a specific policy was implemented. Assuming that this has some effect on case rate, this difference in length of time may play a part in the effect. There are also other external aspects that cannot be included, such as the level of population density broken out among different population demographics.

@@

The covariates that will be examined in building the models fit into two broad categories. The first category is demographic information on the state, namely percentage population distribution of seniors, the unemployment rate, the rate of people living in poverty, and the population density. The second general group is that of policy decisions taken by the state, specifically a mask mandate. Problematic covariates would include any of the other direct measures of COVID-19 severity in the state, namely total infection rate not on a per capita basis and COVID-19 death rates. It can be assumed that these variables would be collinear with the primary variable of interest since COVID-19 case rate is simply a linear transformation of total cases not adjusted for population, and death rate is derived from case rate.

As mentioned in the introduction, the modeling goal for the COVID-19 case rate will be one of description. A model of causation may follow this research if potential causal relationships are examined and incorporated into the model's structure. This may later be expanded to predict the COVID-19 case rate in the past seven days in order for a state to use it to potentially judge what effect a mask policy could have.

The three models aim to investigate the relationship between the COVID-19 case rate, state population demographics, and a mask mandates. COVID-19 case rate per 100,000 is chosen as the key dependent variable to represent the proliferation of the COVID-19 pandemic in each state. The population percentage of seniors is chosen as the key independent variable as a proxy for susceptibility by innate state population make-up. The variables of poverty percentage, unemployment percentage, mask mandate policy, and population density per square miles are chosen as candidate covariates to further explore the aforementioned potential relationship.

@@

```{r message=FALSE}
df %>% 
  ggplot(aes(senior_per_100k, case_rate_100k)) + 
  geom_smooth(se = FALSE) +
  geom_point() +   
  labs(
    title = 'COVID-19 Case Rate due to Senior Population', 
    x = 'Number of seniors per 100,000', 
    y = 'COVID-19 case rate per 100,000'
  )
```

@@

From the QQ plot, normality in the residuals can be detected given the proximity of the points to the normal line. Additionally, although the shape is a little harder to observe, the histogram of the residuals also gives a fairly normal distribution.

@@

```{r message=FALSE}
case_poverty <- df %>% 
  ggplot(aes(poverty_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = ' Association between Covid-19 case rate and the poverty rate', 
    x = 'Number of living under poverty line per 100,000',
    y = 'Covid-19 case rate per 100,000', 
    color = 'Senior num per 100k'
  )

case_unemployed <- df %>% 
  ggplot(aes(unemployed_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Association between Covid-19 case rate and the unemployment rate', 
    x = 'Number of unemployed per 100,000', 
    y = 'Covid-19 case rate per 100,000',
    color = 'Senior num per 100k'
  )

@@

In examining the results of various diagnostics across different models, generally the CLM assumptions are better met as the model continues to develop. In the first model, deviations away from certain assumptions are the strongest. This is seen mostly with regards to the linear conditional expectation assumption, where the residuals against the fitted values have a noticeable inverted "V" shape as opposed to being constant around zero. In the second model, the same diagnostic shows that the assumption is better met. This is further improved upon in the third model, where local variations in the linear conditional expectation diagnostic plot are generally smoother and more centered around zero. The assumption with the greatest exception to this is that of no high colinearity within the models, as each coefficient in each model was shown to have VIF values not near the cut off value of four.

@@

stargazer(model1, model2, model3, type = "text",
          se = list(se.model3),
          star.cutoffs = c(0.05, 0.01, 0.001), 
          title = "Table 1: The effect of population demographics and mask policy on COVID-19 case rate")
```

@@

The goal of the analysis was to address the research question to examine the relationship between both the distribution of population demographics and mask-related policy of a state and the COVID-19 case rate within that states. In order to address this question, the analysis was broken up into three phases of investigation with corresponding linear models:
- Phase 1: To investigate the direct relationship between COVID-19 case rate and the population distribution of seniors within a state
- Phase 2: To build upon Phase 1 by including two additional variables of population demographics, poverty rate and unemployment rate, and investigating the overall effect on COVID-19 case rate
- Phase 3: To build upon Phase 2 by adding a variable to measure the effect of the implementation of a mask-policy on COVID-19 case rate. A variable for population density was also added as an additional measure of population demographics

>>>>

The research question aims to analyze the relationship between the COVID-19 case rate per 100,000 residents within each state and the population demographics features, as well as the policy decisions that were or were not put into place in order to combat the rise of COVID-19 cases. In this sense, the aim is to examine how dependent the COVID-19 case rate is on that of which cannot be controlled (i.e., population demographics), as well as that of which can be controlled to a certain degree (i.e., implementation of policies to attempt to combat case rate) by each state. With this information, one could suggest whether or not the proliferation of COVID-19 within a state seems to be related to either or both controllable and uncontrollable factors and how they differ. The modeling goal of this research question will be one of description, and will be broken up into three phases of investigation.

The first and primary phase of investigation involves how the COVID-19 case rate is related to key population demographics. For this, the key variables are COVID-19 case rate per 100,000 residents and the percentage population distribution of seniors (defined as 65 years old or older in the original dataset) within a state. The case rate per 100,000 was selected as the key output/dependent variable as it provides a standardized measure of the spread of COVID-19 across the states, and already takes into account the absolute population of the states. The senior percentage was selected as the key input/independent variable given that guidance had been released by the Centers for Disease Control and Prevention (CDC) that seniors belong to the age category for those at higher risk of COVID-19, and thus are a well studied and documented group. This is likely due to the both the fact that seniors in this age group are more likely or had greater access to be tested given their categorization and the fact that they would have a greater likelihood of exhibiting detectable COVID-19 related symptoms that would prompt them to get tested. An additional rationale for measuring the senior percentage is that they may also be more likely to contract COVID-19 at lower viral loads, making them more susceptible to the virus. Conversely, this may result in seniors being more wary of the threat of COVID-19, and taking additional precautions to prevent infection relative to other age categories, such as limiting social interactions and taking more preventative measures in terms of hygiene. These key variables will provide an initial understanding of the relationship between the spread of the virus and population demographics.

Following this, in the second phase of investigation the analysis considers other variables of state-level population demographic features that are considered factors of susceptibility, namely the rate of poverty (defined as the percent of individuals living under the federal poverty line in 2018 in the original dataset) and the rate of unemployment (defined as the percent unemployment in 2018 in the original dataset). Those living below the poverty line may have less access/are unable to afford to preventative controls such as sanitation products and masks that help prevent the spread of the virus, while those who are designated as unemployed may not have the flexibility of sheltering at home as well as access to the aforementioned preventative controls. Alternatively, the likelihood of reduced mobility for those living under the poverty line and/or who are unemployed may also play a role in the relationship with case rate. As mentioned, these variables operationalize population metrics that may lead to greater case rates, and are already standardized as rates to account for varying absolute populations across states.

The tertiary and final phase of investigation includes a variable that measures the implementation of policy as a response to COVID-19, and to understand the added effect of this variable in conjunction with the first phase of investigation. Specifically, there is a focus on a policy that mandates the wearing of masks within the state. Through this, the aim is to analyze whether the implementation of a mask-related policy is related to case rate, as well as the strength of the association compared to population metrics. To operationalize this policy and simplify the measurements, the models only consider whether or not this policy was implemented through transformed indicator variables (1 = implemented, 0 = not implemented). As a result, factors of when and for how long the policy was implemented will be lost. Although this loss of information may fail to capture the impact of length of time of a policy on case rate (the policy may take time to be adopted/show meaningful efficacy), as there is no time-series data for case rate included in the data set, it was adjudged to be incongruent with the analysis.

Additionally, the final phase of investigation continues to expand on the examination of state-level population demographics via the population density variable (defined as the population density in square miles, population/square miles in the original dataset). A key factor to the spread of COVID-19 is the idea of social distancing, where the virus is considered to be more likely to spread when people are close in proximity. Thus, population density is measured as a conduit to indicate the level of proximity within each state and will operationalize the concept of social distancing. It should be noted however that the population density data may be limited in its ability to serve as a direct measurement of social distancing given the it lacks other contextualize information, such as the urban to rural ratio as later discussed in section 5.

@@

Secondly, note that there are many internal and external aspects of the selected variables that have not been included within the models. Just addressing the internal information that is lost, it can be seen that some of the information is not captured given the methods of operationalization discussed previously. Among others, the loss of information on the date of implementation within the mask policy variable strips out any contextual knowledge regarding the length of policy implementation, and precludes the identification of how long a specific policy was implemented. Assuming that this has some effect on case rate, this difference in length of time may play a part in the effect. There are also other external aspects that cannot be included, such as the rural to urban ratio as opposed to population density.

@@

The covariates that will be examined in building the models fit into two broad categories. The first category is the demographic features of the state, namely senior rate, the unemployment rate, the poverty rate, and the population density. The second category is that of policy decisions taken by the state, specifically a mask mandate. Problematic covariates would include any of the other direct measures of COVID-19 severity in the state, namely total infection rate not on a per capita basis and COVID-19 death rates. It can be assumed that these variables would be collinear with the primary variable of interest since COVID-19 case rate is simply a linear transformation of total cases not adjusted for population.

The three models aim to investigate the relationship between the COVID-19 case rate, state population demographics, and policy (i.e., mask mandate). COVID-19 case rate per 100,000 is chosen as the key dependent variable to represent the proliferation of the COVID-19 pandemic in each state. The population percentage of seniors is chosen as the key independent variable as a proxy for susceptibility by innate state population make-up. The variables of poverty percentage, unemployment percentage, mask mandate policy, and population density per square miles are chosen as candidate covariates to further explore the aforementioned potential relationship.

@@

```{r message=FALSE}
df %>% 
  ggplot(aes(senior_per_100k, case_rate_100k)) + 
  geom_smooth(se = FALSE) +
  geom_point() +   
  labs(
    title = 'COVID-19 Case Rate vs Senior Population', 
    x = 'Number of seniors per 100,000', 
    y = 'COVID-19 case rate per 100,000'
  )
```

@@

From the Q-Q plot, normality in the residuals can be detected given the proximity of the points to the normal line. Additionally, although the shape is a little harder to observe, the histogram of the residuals also gives a fairly normal distribution.

@@

```{r message=FALSE}
case_poverty <- df %>% 
  ggplot(aes(poverty_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'COVID-19 case rate vs poverty rate', 
    x = 'Number of living under poverty line per 100,000',
    y = 'COVID-19 case rate per 100,000', 
    color = 'Senior num per 100k'
  )

case_unemployed <- df %>% 
  ggplot(aes(unemployed_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'COVID-19 case rate vs unemployment rate', 
    x = 'Number of unemployed per 100,000', 
    y = 'COVID-19 case rate per 100,000',
    color = 'Senior num per 100k'
  )

@@

In examining the results of various diagnostics across different models, generally the CLM assumptions are better met as the model continues to develop. In the first model, deviations away from certain assumptions are the strongest. This is seen mostly with regards to the linear conditional expectation assumption, where the residuals against the fitted values have a noticeable inverted "V" shape as opposed to being constant around zero. In the second model, the same diagnostic shows that the assumption is better met. This is further improved upon in the third model, where local variations in the linear conditional expectation diagnostic plot are generally smoother and more centered around zero. The assumption with the greatest exception to this is that of no high collinearity within the models, as each coefficient in each model was shown to have VIF values not near the cut off value of four.

@@

stargazer(model1, model2, model3, type = "text",
          se = list(se.model3),
          star.cutoffs = c(0.05, 0.01, 0.001), 
          title = paste("Table 1: The effect of population demographics and mask",
          "policy on COVID-19 case rate"))
```

@@

The goal of the analysis was to address the research question to examine the relationship between both the distribution of population demographics and mask-related policy of a state and the COVID-19 case rate within that states. In order to address this question, the analysis was broken up into three phases of investigation with corresponding linear models.

 - Phase 1: To investigate the direct relationship between COVID-19 case rate and the population distribution of seniors within a state
 - Phase 2: To build upon Phase 1 by including two additional variables of population demographics, poverty rate and unemployment rate, and investigating the overall effect on COVID-19 case rate
 - Phase 3: To build upon Phase 2 by adding a variable to measure the effect of the implementation of a mask-policy on COVID-19 case rate. A variable for population density was also added as an additional measure of population demographics
